---
title: 'Time Series HW 2'
authors: 'Blue 6 - Grant, Tae, Manisha, Andrew,Sam'
output: html_document
date: "2023-09-01"
---
## load libraries and data
```{r}
library(tidyverse)
library(lubridate)
library(forecast)
energy = read.csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/AEP_hourly.csv", stringsAsFactors = TRUE)
```

## roll up the data to AVERAGE monthly energy usage
```{r}
energy_df = energy %>%
  mutate(date = mdy_hm(datetime_beginning_ept),
         date = floor_date(date, unit = "month")) %>%
  group_by(date) %>%
  summarise(mavg_mw = mean(mw)) %>%
  arrange(date)
```

## split to train, validation and test data
```{r}
energy_train = energy_df %>%
  filter(date < ymd('2023-1-1'))

energy_valid = energy_df %>%
  filter(date >= ymd('2023-1-1'),
         date <= ymd('2023-4-30'))

energy_test = energy_df %>%
  filter(date > ymd('2023-4-30'))
```

## STL Decomposition
```{r}
mw = ts(energy_train$mavg_mw, start = 2016, frequency = 12)
decomp_stl = stl(mw, s.window = 15)
autoplot(decomp_stl)
```

## variables for decomposition and time series data
```{r}
decomp_df = data.frame(decomp_stl$time.series)

dates = energy_train$date
S_T_R = energy_train$mavg_mw
S_ = decomp_df$seasonal
T_ = decomp_df$trend
```


## Time series and decomposition plot
```{r} 
ggplot() +
  geom_line(aes(x = dates, y = S_T_R, color = "Actual Energy Usage"), lwd = 0.8) +
  geom_line(aes(x = dates, y = T_, color = "Trend Component"), lwd = 0.8) +
  geom_line(aes(x = dates, y = S_, color = "Seasonal Component"), lwd = 0.8) +
  xlab("Time") + ylab("Monthly Average Energy Usage (MW)") +
  scale_color_manual(values = c("Actual Energy Usage" = "black", 
                                "Trend Component" = "blue", 
                                "Seasonal Component" = "red")) +
  guides(color = guide_legend(title = "")) +
  theme(plot.title = element_text(size = 12), 
        axis.title.y = element_text(size = 10), 
        axis.title.x = element_text(size = 10), 
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10), 
        legend.text = element_text(size = 10, face = "plain"),  # Adjust the text size and remove bold
        legend.title = element_text(size = 10))

```


## function to forecast and to calculate MAE and MAPE
```{r}
ESM_Metrics = function(ESM_Model, test_data) {
  test_forecast = forecast(ESM_Model, h = nrow(test_data))

  error = test_data$mavg_mw - test_forecast$mean
  MAPE = mean(abs(error)/abs(test_data$mavg_mw))
  MAE = mean(abs(error))

  return (list(MAPE = MAPE, MAE = MAE))
}
```

## Simple ESM 
```{r}
simple_ESM = ses(mw, initial = 'simple')
ESM_Metrics(simple_ESM, energy_valid)
```
## Holt Trend ESM
```{r}
trend_ESM = holt(mw, initial = 'optimal')
ESM_Metrics(trend_ESM, energy_valid)
```
## Damped Holt Trend ESM
```{r}
trend_ESM_d = holt(mw, initial = 'optimal', damped = TRUE)
ESM_Metrics(trend_ESM_d, energy_valid)
```
## Holt Winters Additive ESM
```{r}
seasonal_ESM_a = hw(mw, seasonal = "additive")
ESM_Metrics(seasonal_ESM_a, energy_valid)
```

## ETS

```{r}
etsModel <- ets(mw)
ESM_Metrics(etsModel, energy_valid)
```

## Holt Winters Additive ESM has the lowest MAPE (0.05734675) - Best Model


## forecast plot on validation data for 'Holt Winters Additive ESM' 
```{r}
valid_forecast = forecast(seasonal_ESM_a, h = nrow(energy_valid))
ggplot() +
  geom_line(aes(x = energy_valid$date, y = energy_valid$mavg_mw, color = "Actual Energy Usage"), lwd = 0.8) +
  geom_line(aes(x = energy_valid$date, y = valid_forecast$mean, color = "Forecasted Energy Usage"), lwd = 0.8) +
  xlab("Time") + ylab("Monthly Average Energy Usage (MW)") +
  guides(color = guide_legend(title = ""))+
  scale_color_manual(values = c("black", "blue"))+
  guides(color = guide_legend(title = "")) +
  theme(plot.title = element_text(size = 12), 
        axis.title.y = element_text(size = 10), 
        axis.title.x = element_text(size = 10), 
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10), 
        legend.text = element_text(size = 10, face = "plain"),  # Adjust the text size and remove bold
        legend.title = element_text(size = 10))
```

## including validation data in the training set
```{r}
energy_train = energy_df %>%
  filter(date <= ymd('2023-4-30'))
```


## Holtwinter additive model on new training data 
```{r}
mw = ts(energy_train$mavg_mw, start = 2016, frequency = 12)

seasonal_ESM_a = hw(mw, seasonal = "additive")
```

## forecasting updated 'Holtwinter additive model' on test data
```{r}
ESM_Metrics(seasonal_ESM_a, energy_test)
test_forecast = forecast(seasonal_ESM_a, h = nrow(energy_test))
```


## forecast plot on training data for 'Holt Winters Additive ESM' 
```{r}
ggplot() +
  geom_line(aes(x = energy_test$date, y = energy_test$mavg_mw, color = "Actual Energy Usage"), lwd = 0.8) +
  geom_line(aes(x = energy_test$date, y = test_forecast$mean, color = "Forecasted Energy Usage"), lwd = 0.8) +
  xlab("Time") + ylab("Monthly Average Energy Usage (MW)") +
  guides(color = guide_legend(title = "")) +
  scale_color_manual(values = c("black", "blue"))+
  theme(plot.title = element_text(size = 12), 
        axis.title.y = element_text(size = 10), 
        axis.title.x = element_text(size = 10), 
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10), 
        legend.text = element_text(size = 10, face = "plain"),  # Adjust the text size and remove bold
        legend.title = element_text(size = 10))
```


