---
title: 'Time Series HW #3'
output: html_document
Author: Andrew,Grant,Manisha,Sam,Tae
date: "09-24-2023"
---
##Load libraries and data
```{r}
library(tidyverse)
library(lubridate)
library(forecast)
energy = read.csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/AEP_hourly.csv", stringsAsFactors = TRUE)
#energy = read.csv("AEP_hourly.csv", stringsAsFactors = TRUE)
```

# rollup data to monthly average
## data split: train, valid, tesr
```{r}
energy_df = energy %>%
  mutate(date = mdy_hm(datetime_beginning_ept),
         date = floor_date(date, unit = "month")) %>%
  group_by(date) %>%
  summarise(mavg_mw = mean(mw)) %>%
  arrange(date)

energy_train = energy_df %>%
  filter(date < ymd('2023-1-1'))

energy_valid = energy_df %>%
  filter(date >= ymd('2023-1-1'),
         date <= ymd('2023-4-30'))

energy_test = energy_df %>%
  filter(date > ymd('2023-4-30'))
```

##STL decomposition
```{r}
mw = ts(energy_train$mavg_mw, start = 2016, frequency = 12)
decomp_stl = stl(mw, s.window = 15)
autoplot(decomp_stl)
```
## Seasonally adjusted data
```{r}
mw_a = mw - decomp_stl$time.series[,1]
autoplot(mw)
autoplot(mw_a)
```
## Dickey fuller test on seasonally adjusted data
## Type 2: with drift and no trend , fail to reject null hypothesis(therfore, random walk with drift)
```{r}
aTSA::adf.test(mw_a)
```

##ACF and PACF plots on differences
```{r}
ggAcf(diff(mw_a, lag = 1)) # spike at lag 1
ggPacf(diff(mw_a, lag = 1)) #spikes at lag 1,2
```
##Fitting ARIMA model ARIMA(2,1,0) and ACF PACF plots on residuals
```{r}
arima_2_1_0 = Arima(mw_a, order = c(2,1,0))
ggAcf(arima_2_1_0$residuals)
ggPacf(arima_2_1_0$residuals)
```
## LjungBox test ( ARIMA(2,1,0)) fail to reject null hypothesis ( no auto correlation)
```{r}
White.LB <- rep(NA, 10)

for(i in 3:10){
White.LB[i] <- Box.test(arima_2_1_0$residuals, 
                        lag=i, type="Ljung-Box", 
                        fitdf = 2)$p.value
}

plot(White.LB, type = "h")
```
## normality of residuals
```{r}
shapiro.test(arima_2_1_0$residuals)
```

##Fitting ARIMA model ARIMA(0,1,1) and ACF PACF plots on residuals
```{r}
arima_0_1_1 = Arima(mw_a, order = c(0,1,1))
ggAcf(arima_0_1_1$residuals)
ggPacf(arima_0_1_1$residuals)
```

## LjungBox test ( ARIMA(0,1,1)) fail to reject null hypothesis ( no auto correlation)
```{r}
White.LB <- rep(NA, 10)

for(i in 2:10){
White.LB[i] <- Box.test(arima_0_1_1$residuals, 
                        lag=i, type="Ljung-Box", 
                        fitdf = 1)$p.value
}

plot(White.LB, type = "h")
```
#normality of residuals ARIMA(0,1,1)
```{r}
shapiro.test(arima_0_1_1$residuals)
```
## forecast function
```{r}
my_forecast = function(model, holdout_data) {
  arima = forecast::forecast(model, h = nrow(holdout_data))$mean
  seasonal = forecast::forecast(decomp_stl$time.series[,1], h = nrow(holdout_data))$mean
  fc = arima + seasonal
  return(fc)
}
```

##metrics function: MAE and MAPE 
```{r}
Metrics = function(forecast, test_data) {
  error = test_data$mavg_mw - forecast
  MAPE = mean(abs(error)/abs(test_data$mavg_mw))
  MAE = mean(abs(error))

  return (list(MAPE = MAPE, MAE = MAE))
}
```

##ARIMA (0,1,1) forecast and metrics
```{r}
F_arima_0_1_1 = my_forecast(arima_0_1_1, energy_valid)
Metrics(F_arima_0_1_1, energy_valid)
```
##Forecast Vs Actual on validation: ARIMA (0,1,1)
```{r}
ggplot() +
  geom_line(aes(x = energy_valid$date, y = energy_valid$mavg_mw), color = "black") +
  geom_line(aes(x = energy_valid$date, y = F_arima_0_1_1), color = "blue") +
  xlab("Date") + ylab("Monthly Average Energy Usage (MW)") +
  ggtitle("Monthly Average Energy Usage (MW) 2016-2023") + ylim(0, 6000)
```
##ARIMA (2,1,0) forecast and metrics
```{r}
F_arima_2_1_0 = my_forecast(arima_2_1_0, energy_valid)
Metrics(F_arima_2_1_0, energy_valid)
```
##Forecast Vs Actual on validation: ARIMA (2,1,0)
```{r}
ggplot() +
  geom_line(aes(x = energy_valid$date, y = energy_valid$mavg_mw), color = "black") +
  geom_line(aes(x = energy_valid$date, y = F_arima_2_1_0), color = "red") +
  xlab("Date") + ylab("Monthly Average Energy Usage (MW)") +
  ggtitle("Monthly Average Energy Usage (MW) 2016-2023") + ylim(0, 6000)
```
##PLots: Forecast Vs Actual on validation: ARIMA (0,1,1) and ARIMA(2,1,0)
```{r}
ggplot() +
  geom_line(aes(x = energy_df$date[1:84], y = energy_df$mavg_mw[1:84]), color = "black") +
  geom_line(aes(x = energy_valid$date, y = F_arima_0_1_1), color = "blue") +
  geom_line(aes(x = energy_valid$date, y = F_arima_2_1_0), color = "red") +
  xlab("Date") + ylab("Monthly Average Energy Usage (MW)") +
  ggtitle("Monthly Average Energy Usage (MW) 2016-2023") + ylim(0, 6000)
```

# Adding Validation data into training
```{r}
energy_df = energy %>%
  mutate(date = mdy_hm(datetime_beginning_ept),
         date = floor_date(date, unit = "month")) %>%
  group_by(date) %>%
  summarise(mavg_mw = mean(mw)) %>%
  arrange(date)

energy_train = energy_df %>%
  filter(date <= ymd('2023-4-30'))

energy_test = energy_df %>%
  filter(date > ymd('2023-4-30'))
```

##STL decomposition on new train data
```{r}
mw = ts(energy_train$mavg_mw, start = 2016, frequency = 12)
decomp_stl = stl(mw, s.window = 15)
```

##Seasonally adjusted on new train data
```{r}
mw_a = mw - decomp_stl$time.series[,1]
autoplot(mw)
autoplot(mw_a)
autoplot(diff(mw_a, lag = 1))
```
##Fitting ARIMA(2,1,0) and ARIMA(0,1,1) on new train data
```{r}
arima_2_1_0 = Arima(mw_a, order = c(2,1,0))
arima_0_1_1 = Arima(mw_a, order = c(0,1,1))
```

##forecast on test data ARIMA(0,1,1)
```{r}
F_arima_0_1_1 = my_forecast(arima_0_1_1, energy_test)
Metrics(F_arima_0_1_1, energy_test)
```
##forecast on test data ARIMA(2,1,0)
```{r}
F_arima_2_1_0 = my_forecast(arima_2_1_0, energy_test)
Metrics(F_arima_2_1_0, energy_test)
```
##forecast plot of ARIMA (0,1,1)
```{r}
ggplot() +
  geom_line(aes(x = energy_test$date, y = energy_test$mavg_mw, color = "Actual Energy Usage"), size = 1) +
  geom_line(aes(x = energy_test$date, y = F_arima_0_1_1, color = "Forecasted Energy Usage"), size = 1) +
  xlab("Date") + ylab("Monthly Average Energy Usage (MW)") + ylim(0, 6000) +
  scale_color_manual(values = c("Actual Energy Usage" = "black", "Forecasted Energy Usage" = "blue")) +
  labs(color = "") 
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.title.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10, face = "plain"),  # Adjust the text size and remove bold
        legend.title = element_text(size = 10),
        legend.position = "right")  # Adjust legend position
```

##forecast plot of ARIMA(2,1,0)
```{r}
ggplot() +
  geom_line(aes(x = energy_test$date, y = energy_test$mavg_mw, color = "Actual Energy Usage"), size = 1) +
  geom_line(aes(x = energy_test$date, y = F_arima_2_1_0, color = "Forecasted Energy Usage"), size = 1) +
  xlab("Date") + ylab("Monthly Average Energy Usage (MW)") + ylim(0, 6000) +
  scale_color_manual(values = c("Actual Energy Usage" = "black", "Forecasted Energy Usage" = "red")) +
  labs(color = "") 
  theme(plot.title = element_text(size = 12, hjust = 0.5),
        axis.title.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10, face = "plain"),  # Adjust the text size and remove bold
        legend.title = element_text(size = 10),
        legend.position = "right")  # Adjust legend position
```

##forecast plot of ARIMA (0,1,1) and ARIMA(2,1,0)
```{r}
ggplot() +
  geom_line(aes(x = energy_df$date[1:88], y = energy_df$mavg_mw[1:88]), color = "black") +
  geom_line(aes(x = energy_test$date, y = F_arima_0_1_1), color = "blue") +
  geom_line(aes(x = energy_test$date, y = F_arima_2_1_0), color = "red") +
  xlab("Date") + ylab("Monthly Average Energy Usage (MW)") +
  ggtitle("Model2 - Monthly Average Energy Usage (MW) 2016-2023") + ylim(0, 6000) + theme(plot.title = element_text(hjust = 0.5))

```

